Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
new text for dynamic eval

\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{OOPSLA} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\setcopyright{none}
\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{my_style}
\usepackage{listings, wrapfig,xspace}
\usepackage{paralist}

@@

\newcommand{\NUMFUNCTIONS}{25,000\xspace}  %%% TODO auto-gen
\newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\PACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\genthat}{genthat\xspace}  %%% TODO auto-gen
\newcommand{\YEARS}{20\xspace} %%TODO fix

@@

This paper focuses on a simpler problem than an entire type system, instead,
we limit the scope of our investigation to giving types to function
signatures. In order to do this, we design a simple type language, one that
matches the R data types but omits features such as parametric polymorphism
and subtyping between user-defined data types. We then extract type
signature from execution traces of a corpus of widely used libraries.  This
allows us to see how far one can get with the simple type language and
identify limitations of our this particular design. We validate the
robustness of the extracted type signature by implementing a contact system
that weaves the types around their respective functions, and use a large
number of clients of the target packages for validation. The contract system
can be used by the R community to experiment with our type signatures and to
replace ad hoc error checking code.

To sum up, our paper makes the following main contributions:
\begin{itemize}
\item We implemented tooling to automatically extract type signatures from R
  functions and to intrument R functions with checks based on their declared
  types. The tooling is robust and scalable to the entire R language.
\item We carried out a large-scale analyis of corpus of \PACKAGES widely
  used and actively maintained libraries to extract function type signatures
  validated the robustness of the inferred type signatures against XXX
  programs that use those functions.
\item We report on the appropriateness and usefulness of a simpe type
  language for the R programming language.
\end{itemize}

Our tools are open source and publicly available on GitHub (link anonymize),
all results in this paper are reproducible and will be submitted for
artifact evaluation should this paper be accepted.
%
\section{Background} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

As mentioned, our analysis is built on the R-dyntrace framework.  R-dyntrace
is a very efficient instrumentation framework for R, which is built on a
modified version of the R runtime.

@@

@@

%
%
%
\subsection{contractr}
\label{sec:contractr}

@@

%
%
%
\section{Type Language Design}
\label{sec:typesystemdesign}

@@

A common primitive usage pattern is related to the presence of NAs:
programmers explicitly check for NAs in primitive vectors, possibly
sanitizing them if they are present.  NAs complicate practically every
computation using primitive vectors, a fact that many programmers appear to
be acutely aware of. For instance, consider the code in
Figure~\ref{fig:na-example}.

@@

This code is from later in the same \code{rowWeightedMeans} function
discussed previously.  Here, the length of the weights vector \code{w} is
being checked for equality with the number of columns in the matrix
\code{x}.  Specifying this relationship as a type amounts to {\it dependent
  typing}, which we believe too complex to retrofit onto a dynamic data
science language such as R.  Ultimately, we decided to leave these kinds of
checks up to the programmer.

@@

We are faced with several possible designs for list types.  Chief concerns
include the heterogeneity of list elements and the presence list element
names, which allow list elements to be accessed in a struct-like manner.
The next sections highlight some important design decisions for our list
types.

@@

@@

%
\subsubsection{Data Frames}

@@

It will also require proper treatment of function types, and perhaps even
parametric types, which dynamic analyses such as \typetracer are unable to
discern soundly.

@@

@@

We have discussed most of these types throughout this section, with some
notable exceptions.  ${\bf expression}$, ${\bf language}$, ${\bf symbol}$,
${\bf externalptr}$, ${\bf pairlist}$, ${\bf S4}$, and ${\bf weakref}$ types
occur \AT{rarely}, and are by-and-large uninteresting to most R programmers.
\AT{Fill in details of these basic types.}  One other type we did not
discuss is the union type $T_1 \, | \, T_2$, which is supported by
\contractr and generated by \typetracer when it cannot find some way to
consolidate argument types (for example, if an argument is observed to be
${\bf chr}$ and ${\bf int}$, \typetracer will ascribe type ${\bf chr} \, |
\, {\bf int}$ to it).  And while not generated by \typetracer (recall
discussion in Section~\ref{subsec:structs}), user-defined struct types have
the form ${\bf struct}\langle F_1, ..., F_n \rangle$, with fields $F_i$
being name $:$ type pairs.

@@

%
%
\subsubsection{Results}

\AT{How many failing assertions? How many functions had these failed assertions? Which packages? How was the coverage? What are patterns in the actual types of the failed assertions? In the expected types? How many times did the code trigger the contracts?}

Overall, we found that only \AT{2\%} of contract assertions failed.
This indicates two things: firstly, the corpus of packages we selected is quite well-tested, and much of the intended function behaviour is exemplified in the package example code. 
Secondly, and more importantly, this success indicates that our analysis framework is able to generate function types which accurately capture the behaviour of the function. 
Of the \AT{2\%} of failing contract assertions, we picked up on a few significant patterns.

Firstly, we found that \AT{10\%} of the failing assertions were due to \code{double}-typed values being passed to function arguments expecting an \code{integer}.
While we cannot say for sure that {\it all} of these cases could be handled by coercion, it is worth noting that coercion from doubles to integers is exceedingly common in R: the practice is not codified in the language spec, but many of R's internal C functions explicitly perform this coercion.
\AT{We can put an example, e.g. nrow(df) is an integer but 5 is not. Also list access coerces the double to an int.}

Next, \AT{30\%} of failing assertions were due to vector-typed primitive values being passed to function arguments expecting a \code{class<matrix>}.
This case exemplifies something we have touched on earlier, namely that vectors can be readily considered to be unidimensional matrices.
\AT{Is this coercion automatic? I've seen it mentioned often in the documentation.}

Of the remaining \~60\% of failed contract assertions, we found quite an interesting pattern: the top as-of-yet unexplained contract failure, making up roughly 7\% of total failed assertions, is an assertion checking a ${\bf character}$ against the expected type ${\bf class} \langle bignum \rangle \, | \, {\bf raw}[]$.
Further investigation revealed that most of these failures were coming from the \code{openssl::base64_encode} function, whose code is available in Figure~\ref{fig:base64-encode}.

\begin{lstlisting}
# The type according to typetracer:
#' @type <bin: class<`bignum`> | raw[], linebreaks: logical> => character
base64_encode <- function (bin, linebreaks = FALSE) 
{
    if (is.character(bin)) {
        bin <- charToRaw(paste(bin, collapse = ""))
    }
    stopifnot(is.raw(bin))
    .Call(R_base64_encode, bin, as.logical(linebreaks))
}
\end{lstlisting}

\caption{Example of reverse dependency increasing code coverage.}
\label{fig:base64-encode}
\end{center}
\end{figure}

Contract failures are on the first argument, \code{bin}.
This function is a wrapper for a C function (\code{R_base64_encode}) which encodes the input \code{bin} in base64.
It would appear that the code for this function was never tested with a ${\bf character}$-typed input, even though it {\it is} valid.
This is an example of a rather common phenomenon wherein the code of a package's reverse dependencies increases the coverage of some of the original package's functions, exemplified here by the code on line 6 never having been run in \code{openssl}'s own tests.

To try to detect this pattern, we cross-referenced functions with failing contract checks against the code coverage for that function from the package's tests.
For a given function, a coverage of X\% indicates that X\% of the code in the function body was executing by running package test, example, and vignette code.
Of the remaining failed contracts, we found that \~28\% of them were from functions which had less than 100\% coverage, meaning that it is possible that \typetracer failed to synthesize the correct type for the function as not all possible executions were recorded.

But what of the large amount (\~70\%) of assertion failures in functions with 100\% coverage?
As it happens, many of these functions have very few lines of code, many having as few as a single line.
The most interesting among these are functions which are the entry point for S3 dispatch (these functions typically have a single line of code, \code{UseMethod("InsertS3MethodNameHere")}).

It turns out that the lion's share of the remaining polymorphism is related to S3 dispatch.
S3 generics are naturally polymorphic, and we aim to define types to capture these cases, and implement a full-fledged object-oriented type system for R in future work.
In the meantime, if we count out S3 methods, we find that only \AT{few} contract failures remain.

>>>>

\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{OOPSLA} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\setcopyright{none}
\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{my_style}
\usepackage{listings, wrapfig,xspace}
\usepackage{paralist}
\usepackage{booktabs} % To thicken table lines

@@

\newcommand{\NUMFUNCTIONS}{25,000\xspace}  %%% TODO auto-gen
\newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\PACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\genthat}{genthat\xspace}  %%% TODO auto-gen
\newcommand{\YEARS}{20\xspace} %%TODO fix
\newcommand{\PERCFAILEDASSERTIONS}{2\%\xspace}
\newcommand{\NUMPKGSEVAL}{7000\xspace} % autogenerate
\newcommand{\NUMASSERTIONS}{39210391\xspace} %autogenerate
\newcommand{\PROPFUNSFAILEDCHECK}{21.21\%\xspace}
\newcommand{\PROPFUNSFAILEDCHECKNOSTHREE}{15.48\%\xspace} 
\newcommand{\PROPARGSFAILINGASSERTS}{3.53\%\xspace} % autogen....

@@

In order to do this, we need to understand the degree of polymorphism
present in R code, that is to say, how programmers leverage the dynamic
nature of R to write code that can accept arguments of different types.
This understanding will drive our design.

We propose to capture the degree of polymorphism present in R by means
of a dynamic analysis of widely used libraries. For each function call we
can record the types of its arguments and of its return value. This allows
us to observe how many different combination of types are accepted by any
given function. Unlike many other languages, R has a carefully curated
software repository called CRAN. To be deposited in CRAN, a package must
come with sample dataset, tests and executable use-cases. As part of normal
operations these tests are run regularly and failing packages are removed.
This allowed us to have access to \PACKAGES libraries and about an order of
magnitude more runnable scripts that exercise those libraries.

The contributions of this paper are thus as follows:
\begin{itemize}
\item A large-scale analysis of the polymorphism present in function
  signatures of \PACKAGES widely used and actively maintained R packages.
\item A tracing and analysis pipeline that extends a previously published
  test generation tool named \genthat.
\item A set of type annotations that captures how programmers use types
in existing R code.
\end{itemize}

One threat to validity of our work is that we rely on dynamic analysis, so
our conclusions are only as good as the coverage of the possibly function
behaviors. Previous work~\cite{issta18} reported that running all the
scripts that come with CRAN packages gives, on average, 68\% test coverage.
We attempted to mitigate the threat coming from the fact that only part of
the code is being exercised by manual analysis. It would be reasonable to
ask for confirmation of the data by static analysis of the code, but sound
static analysis of R is difficult because of the extensive use of reflective
features such as \code{eval} and of the ability to redefine the meaning of
operators such as \code{+} and \code{if}.  Another threat to validity is
that we only have access to code that has been deposited in the CRAN
repository. While this may bias our findings towards code written to be
reusable and, possibly, better engineered than typical user code. This is
also the code that would most benefit from type annotations.

%
\section{Background} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

As mentioned, our analysis is built on the R-dyntrace framework.  R-dyntrace
is a very efficient instrumentation framework for R, which is built on a
modified version of the R runtime. 
% Our dynamic type analysis only needs to be run once, ever---from the analysis results, we synthesize function signatures which are passed along to later type-checking tools which do not rely on a heavily modified runtime.

@@

\AT{Maybe further discuss how we can create many candidate TS from the data.}

@@

%
%
%
%
\subsection{contractr}
\label{sec:contractr}

@@

%
%
%
%
%
%
\section{Type Language Design}
\label{sec:typesystemdesign}

@@

A common primitive usage pattern is related to the presence of NAs:
\AT{often}, programmers explicitly check for NAs in primitive vectors,
possibly sanitizing them if they are present.  NAs complicate practically every computation
using primitive vectors, a fact that many programmers appear to be acutely
aware of. For instance, consider the code in Figure~\ref{fig:na-example}.

@@

This code is from later in the same \code{rowWeightedMeans} function discussed previously.
Here, the length of the weights vector \code{w} is being checked for equality with the number of columns in the matrix \code{x}. 
Specifying this relationship as a type amounts to {\it dependent typing}, which we believe too complex to retrofit onto a dynamic data science language such as R.  
Ultimately, we decided to leave these kinds of checks up to the programmer.
\AT{Can we say more here?}

@@

We are faced with several possible designs for list types.  
Chief concerns include the heterogeneity of list elements and the presence list element names, which allow list elements to be accessed in a struct-like manner.  
% \AT{Rarely?}, programmers even write code expecting lists of a particular pre-defined length.  
The next sections highlight some important design decisions for our list types.

@@

%
%
% \subsubsection{Forgoing Tuples, or Short, Constant Length Lists}

% We briefly entertained the possibility of including a \code{tuple} type, which would have looked like e.g. \code{tuple<int, int>} for a list of two integers.  
% When configured to detect tuples, our analysis did indeed pick up \AT{a number}, but post-processing and simplification revealed that in \AT{many} cases, they would be subsumed by co-occurring \code{list} types.
% In other words, more often than not it just so happened that functions were being called with short lists.  
% Of the cases that tuples were not subsumed by lists, \AT{often} there were \AT{tuples of different lengths}, implying that the shape of the argument was not essential.

% \AT{I feel like this is the least important of the design decisions we made, so I'm leaving it for the end, we might not even want it.}

@@

% We also thought to encode data frames as structs, as they are essentially lists with named columns of homogeneous values, but it was difficult to draw any meaningful conclusions from the data as unfortunately structs introduced a lot of noise into our analysis.

%
%
\subsubsection{Data Frames}

@@

It will also require proper treatment of function types, and perhaps even parametric types, which dynamic analyses such as \typetracer are unable to discern soundly.
\AT{General words to convince the reader that data frames are, in fact, hard, and worthy of their own paper.}

@@

\AT{Revisit this section since discussion of polymorphism in results.}

@@

We have discussed most of these types throughout this section, with some notable exceptions.
${\bf expression}$, ${\bf language}$, ${\bf symbol}$, ${\bf externalptr}$, ${\bf pairlist}$, ${\bf S4}$, and ${\bf weakref}$ types occur \AT{rarely}, and are by-and-large uninteresting to most R programmers.
\AT{Fill in details of these basic types.}
% $S$ and $V$ are scalar and vector primitives, with vectors taking the form $S[]$.
% The scalar type $S$ can be either a base type $B$ (one of the 6 primitives in R), or an NA-able base type (roughly, $\string^B$).
One other type we did not discuss is the union type $T_1 \, | \, T_2$, which is supported by \contractr and generated by \typetracer when it cannot find some way to consolidate argument types (for example, if an argument is observed to be ${\bf chr}$ and ${\bf int}$, \typetracer will ascribe type ${\bf chr} \, | \, {\bf int}$ to it).
And while not generated by \typetracer (recall discussion in Section~\ref{subsec:structs}), user-defined struct types have the form ${\bf struct}\langle F_1, ..., F_n \rangle$, with fields $F_i$ being name $:$ type pairs.

@@

%
%
\subsubsection{Breakdown of Contract Assertion Failures}

We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS total assertions.
Overall, we found that only \PERCFAILEDASSERTIONS of contract assertions failed.
This number is low, and suggests that the type signatures generated by running \typetracer on our corpus are indeed representative, and capture intended function behaviour.

We breakdown the \PERCFAILEDASSERTIONS of failed assertions by type in Table~\ref{tbl:fail-breakdown-0}.
By far the most commonly failed assertion is one asserting that a value with ${\bf dbl}[]$ type has type ${\bf class}\langle matrix \rangle$.
This was a deliberate design decision, as coercion from vectors to matrices is ad hoc at best, and surely not a practice codified in the language.
In a similar vein, another popular failing assertion is checking if a ${\bf dbl}[]$ has type ${bf int}[]$, another case of commonly performed coercion.
We did not include these types of coercions in our type annotation framework as programmers cannot rely on them, and it is not always the case that the coercions are safe to perform.

\begin{table}
% BEGIN Autogenerated

\begin{tabular}{llrrr}
\toprule
Actual Type & Expected Type & \# Failed Assertions & \% Total & Cumulative \%\\
\midrule
double[] & class<`matrix`> & 371896 & 27.37 & 27.37\\
double[] & integer[] & 117212 & 8.62 & 35.99\\
character & class<`bignum`> | raw[] & 100100 & 7.37 & 43.36\\
double & (integer | null) & 60570 & 4.46 & 47.81\\
double[] & class<`timeSeries`> & 58872 & 4.33 & 52.15\\
\addlinespace
class<matrix> & class<`timeSeries`> & 55005 & 4.05 & 56.19\\
double[] & double & 53338 & 3.92 & 60.12\\
double & class<`data.frame`> & 32553 & 2.40 & 62.51\\
double[] & class<`data.frame`> & 31576 & 2.32 & 64.84\\
double & integer & 20586 & 1.51 & 66.35\\
\bottomrule
\end{tabular}
% END Autogenerated
\label{tbl:fail-breakdown-0}
\caption{Top types of contract assertion failures.}
\end{table}

In addition to the raw number of failing contract checks, we were interested in how many functions had a parameter where a contract check failed.
Overall, we found that \PROPFUNSFAILEDCHECK of functions had a failing contract check.
Digging a little deeper, we noticed that many of these functions were performing S3 dispatch, which is part of R's object-orientation framework which we will tackle in future work.
Removing those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.

Turning our attention now to arguments, we found that only \PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
This is promising, as by-and-large it appears that \typetracer inferred argument type signatures which were not violated during testing.
But what do these failing assertions look like?
Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting runtime occurrences, but that data alone does not tell the full story, as some failures may be overrepresented if e.g. a failing contract assertion was in a loop.
Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by parameter position, folding away multiple identical failed contract assertions for the same parameter position.
The first row of this table reads: a value of type ${\bf chr}[]$ was passed to 183 different argument positions expecting a value of type ${\bf chr}$, representing 2.05\% of all arguments where a contract check failed.

The main takeaway from Table~\ref{tbl:fail-breakdown-1} is that there appear to be no pathological kinds of failing assertions: programmers are not disproportionately passing values of one type to arguments expecting values of another.
\AT{I need to touch up the discussion of this table.}

% While the raw number of failing checks is interesting, we were also curious to see if certain kinds of type checks were overrepresented in the data.
% A breakdown of the top failing type checks by kind can be found in Table~\ref{tbl:fail-breakdown-2}.
% % Note the last row in the table: this is a check that would succeed in a proper type system with subtyping for class types, as data frames are subclasses of tibbles (\code{tbl}), and we will revisit these scenarios in future work.

 \begin{table}
% BEGIN Autogenerated

 \begin{tabular}{llrrr}
\toprule
Actual Type & Expected Type & \# Arguments & \% Total & Cumulative \%\\
\midrule
 character[] & character & 183 & 2.05 & 2.05\\
 \textasciicircum{}logical & null & 158 & 1.77 & 3.81\\
 character & null & 117 & 1.31 & 5.12\\
 double & null & 107 & 1.20 & 6.32\\
 double[] & double & 100 & 1.12 & 7.44\\
 class<data.frame> & ${\bf class} \langle data.frame, tbl, tbl\_df \rangle$ & 63 & 0.70 & 8.14\\
 \bottomrule
\end{tabular}

% END Autogenerated
 \label{tbl:fail-breakdown-1}
 \caption{Most common failing contract checks. Note that for this table, we collapsed multiple identical failed assertions into one, meaning that single arguments with multiple (identical) failed assertions are only counted once.}
 \end{table}
